The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-23 dictionary.tex
Changes to Introduction and Applications
Modified comments regarding Bernstein; fixed the proposition using Weil
restriction and Néron models; changed the abstract and title.

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes over finite fields,
  in which case some interesting new features of the dictionary arise.
   \end{abstract}

@@

Motivated by an application to the geometrization and categorification of quasicharacters of $p$-adic tori,
this paper extends Deligne's function--sheaf dictionary for characters of connected commutative algebraic groups over finite fields to characters of smooth commutative group schemes over finite fields.

@@

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(\Gr_R(X))$ of quasicharacter sheaves on $\Gr_R(X)$.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the Néron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

Specializing to the case when $X_K$ is an algebraic tori $T$ over $K$, and writing $\underline{T}$ for the Neron model $X$ in this case, this paper gives a canonical short exact sequence 
\begin{equation}
0 \to \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCiso{T} \to \Hom_\text{qc}(T(K),\EEx) \to 0,
\end{equation}
where $\Hom_\text{qc}(T(K),\EEx)$ denoted the group of quasicharacters of $T(K)$.
We further show that automorphism groups in the $\QC(\Gr_R(\underline{T}))$ are given, for every quasicharacter sheaf $\qcs{F}$ on $\Gr_R(X)$, by
\begin{equation}
\Aut(\qcs{F}) \iso (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.
We also use quasicharacter sheaves to categorify the transfer of quasicharacters of tori between characteristic $0$ and characteristic $p$ local fields.

Joseph Bernstein has suggested to us that, although quasicharacter sheaves on $\Gr(\underline{T})$ do not correspond exactly to quasicharacters of $T(K)$, the category $\QC(\Gr(\underline{T}))$ may be closely tied to characters of the pure rational forms of $T$ in the sense of Vogan \cite{vogan:93a} or perhaps the rigid inner twists of Tasho Kaletha. 
A better understanding of 
$\Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{}}$ should provide insight into this suggestion. \todo{rephrase}

@@

%\begin{acknowledgements}
\subsection{Acknowledgments}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who made some very helpful observations and suggestions.
%
Finally, we gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our week-long stay in May 2012.

@@

We will describe the group $\QCiso{G}$ in Theorem \ref{thm:snake}
and the sets $\Hom(\qcs{L},\qcs{L}')$ in Theorem \ref{thm:autornaught}; in this way we provide a complete description of the category $\QC(G)$.

@@

@@

Let $(\gqcs{L},\phi)$ be a Weil sheaf on $G$. Every $g\in G(\Fq)$
determines a point $\bg$ on $\bG$ fixed by $\Frob{G}$. 
Together with the canonical isomorphism $(\Frob{G}^*\gqcs{L})_{\bg} \iso  \gqcs{L}_{\Frob{G}(\bg)}$, in this way the automorphism $\phi$ determines an automorphism $\phi_{\bg}$ of the $\EE$-vector space $\gqcs{L}_{\bg}$; let $\trace(\phi_{\bg};\gqcs{L}_{\bg})$ be the trace of $\phi_{\bg} \in \Aut_{\EE}(\gqcs{L}_{\bg})$ and let $\trFrob{(\gqcs{L},\phi)} : G(\Fq)\to \EE$ be the function defined by 
\begin{equation}\label{trWeil}
%\begin{equation}\label{trWeil}
\trFrob{(\gqcs{L},\phi)}(g) \ceq \trace(\phi_{\bg};\gqcs{L}_{\bg}),
%\end{equation}
\end{equation}
commonly called the {\em trace of Frobenius of $(\gqcs{L},\phi)$}.
Note that if $(\gqcs{L},\phi) \iso (\gqcs{L'},\phi')$ as Weil sheaves, 
then $\trFrob{(\gqcs{L},\phi)} = \trFrob{(\gqcs{L'},\phi')}$ as functions on $G(\Fq)$.

@@

\begin{proposition}\label{prop:pullback}
  If $f : H\to G$ is a morphism of smooth commutative group schemes over $\Fq$ then
  \begin{align*}
  f^* : \QC(G) &\to \QC(H) \\
  (\gqcs{L},\mu,\phi) &\mapsto (\bf^*\gqcs{L},(\bfxf)^*\mu,\bf^*F)
  \end{align*}
  defines a monoidal functor dual to $f \colon H(\Fq) \to G(\Fq)$:
  \[
  \begin{tikzcd}[row sep=20, column sep=30]
   \QCiso{G} \rar{f^*} \arrow[swap]{d}{\TrFrob{G}} & \QCiso{H} \dar{\TrFrob{H}} \\
   G(\Fq)^* \rar & H(\Fq)^*
  \end{tikzcd}
  \]
  is a commutative diagram of groups.  Moreover, $(f\circ g)^* = g^* \circ f^*$.
\end{proposition}

\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. 
  Pullback by $\bf$ takes rank~one local systems to rank~one local systems.
  To see that $(\bfxf)^* \mu$ satisfies \ref{QC.2},
  apply the functor $(\bfxf)^*$
  to \ref{QC.2} for $\qcs{L}$ and use the canonical isomorphism
  $(\bfxf)^*(\LxL) \iso \bf^*\gqcs{L} \tight{-3}{\boxtimes}{-3} \bf^*\gqcs{L}$.
  To show that $f^*\qcs{L}$ satisfies
  \ref{QC.3}, apply the same functor to \ref{QC.3} for $\qcs{L}$.
  Since $f$ is a morphism of group schemes defined over $\Fq$
  it provides isomorphisms $(\bfxf)^*\Frob{\GxxG}^* \iso \Frob{\GxxG}^* (\bfxf)^*$
  and $(\bfxf)^* \bm^*\iso \bm^* \bf^*$ between functors of constructible sheaves.

@@

\begin{proposition}\label{prop:product}
If $G_1$ and $G_2$ are smooth commutative group schemes over $\Fq$ then the rule
$\boxtimes : (\qcs{L}_1,\qcs{L}_2)\to \qcs{L}_1\boxtimes\qcs{L}_2$ defines an equivalence of categories
\[
\QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)
\]
such that
\[
\begin{tikzcd}[column sep=60]
\arrow{d}{\TrFrob{G_1} \times \TrFrob{G_2}} \QCiso{G_1}\times \QCiso{G_2} \arrow{r}{\boxtimes}
& \arrow{d}{\TrFrob{G_1\times G_2}} \QCiso{G_1\times G_2}\\
(G_1)(\Fq)^*\times (G_2)(\Fq)^* \arrow{r}{(\chi_1,\chi_2)\mapsto \chi_1\otimes \chi_2}  & (G_1\times G_2)(\Fq)^*
\end{tikzcd}
\]
commutes.
\end{proposition}

\begin{proof}
The only non-trivial part of the proof is to show that $\boxtimes : \QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)$
is essentially surjective, so we will only address that point here.

@@

Now, let $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
Set $\qcs{L}_1 \ceq q_1^* \qcs{L}$ and $\qcs{L}_2 \ceq q_2^* \qcs{L}$;
by Proposition~\ref{prop:pullback}, $\qcs{L}_1$ a quasicharacter sheaf on $G_1$
and $\qcs{L}_2$ is a quasicharacter sheaf on $G_2$.
We will obtain an isomorphism $\qcs{L} \iso  \qcs{L}_1\boxtimes \qcs{L}_2$.
Applying the functor $f^*$ to the isomorphism $\mu$ yields
\begin{equation}\label{eq:fm}
f^*\mu : f^* m^* \gqcs{L} \to f^*(\gqcs{L}\boxtimes \gqcs{L}) .
\end{equation}
We have already observed that $m\circ f = \id_G$, so $f^* m^* \gqcs{L} = \gqcs{L}$.
On the other hand, $f^*(\gqcs{L}\boxtimes \gqcs{L}) = f^*p_1^*\gqcs{L}\otimes f^* p_2^*\gqcs{L}$
by definition.  Since $f^*p_1^*\gqcs{L} = r_1^* q_1^* \gqcs{L}$ and $f^*p_2^*\gqcs{L} = r_2^* q_2^* \gqcs{L}$
we get that $f^*(\gqcs{L}\boxtimes \gqcs{L}) = \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It follows that \eqref{eq:fm} gives an isomorphism $\gqcs{L} \to  \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It is a matter of routine to show that this morphism satisfies
\ref{QC.3} as it applies here,
from which it follows that we have exhibited an isomorphism
$\qcs{L} \to \qcs{L}_1\boxtimes \qcs{L}_2$ of quasicharacters sheaves on $G\times G$.
This completes the proof that $\boxtimes$ is essentially surjective.
\end{proof}

@@

\begin{proof}
The first part of Proposition~\ref{prop:pullback} shows that $F_1$ is a functor,
while the second part shows that Trace of Frobenius is a natural transformation
$T: F_1 \to F_2$. When further combined with Proposition~\ref{prop:product},
we see that $F_1$ is an additive functor and $T: F_1 \to F_2$ is a natural
transformation between additive functors,
concluding the proof of Proposition~\ref{prop:functorialG}.
\end{proof}

@@

\subsection{Stalks of quasicharacter sheaves}

@@

\begin{proposition}
\begin{enumerate}
\item The second page is given by $E_2^{i,j} = \Hh^i(\Weil{}, \Hh^j(G(\bFq), \EEx))$,
\item there is an isomorphism $\Hh^n(\Weil{} \ltimes G(\bFq), \EEx) \cong \Hh^n(E_G^\bullet)$,
\item there is a filtration $\Hh^n(\Weil{} \ltimes G(\bFq), \EEx) = F_n \supset \cdots \supset F_0 \supset F_{-1} = 0$ where $F_i / F_{i-1} \cong E_{\infty}^{i, n-i}$.
\end{enumerate}
\end{proposition}

@@

\begin{proof}
Suppose $\rho \in \Hom(\qcs{L},\qcs{L}')$.
If $\rho_0 = 0$ then the second diagram in \ref{qc.4} for $y=0$ implies that $\rho_x = 0$ for all $x\in G$, in which case $\rho$ is trivial.
On the other hand, if $\rho_0$ is nonzero then the second diagram in \ref{qc.4} for $y = -x$ implies that $\rho_x$ is nonzero for all $x$ and thus an isomorphism since the stalk of quasicharacter sheaves are one-dimensional; in this case $\rho$ is an isomorphism.

Now suppose $\rho \in \Aut(\qcs{L})$.
The second diagram in \ref{qc.4} shows that the association $x \mapsto \rho_x$ is a homomorphism from $G(\bFq)$ to $\EEx$ and the first diagram in \ref{qc.4} shows that it factors through $G(\bFq)_{\Weil{}}$.  Conversely, if $\rho : G(\bFq)_{\Weil{}} \to \EEx$ is any homomorphism then defining $\rho_x$ as multiplication by $\rho(x)$ will satisfy the two diagrams.  Composition of morphisms corresponds to pointwise multiplication in this correspondence, showing that the resulting bijection is actually a group isomorphism.
\end{proof}

@@

\begin{proposition}\label{prop:connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then:
 \begin{enumerate}
 \item $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is an isomorphism of groups;
 \item every quasicharacter sheaf on $G$ is isomorphic to one defined by a discrete isogeny;%as in Proposition~\ref{prop:finite};
 \item $\Aut(\qcs{L}) = 1$, for all quasicharacter sheaves $\qcs{L}$ on $G$;
  \item for all quasicharacter sheaves $\qcs{L}$, $\qcs{L}'$ on $G$ and for every $\rho \in \Hom(\qcs{L},\qcs{L}')$, either $\rho$ is trivial or an isomorphism.
 \end{enumerate}
 \end{proposition}

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by étale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
For the last part of the proposition, argue as in the analogous part of Proposition~\ref{prop:autornaught_etale}.
\end{proof}

@@

\section{The function--sheaf dictionary for smooth, commutative group schemes}\label{sec:main}

@@

\begin{proof}
The sequence is exact at $\QCiso{G^0}$ by Proposition~\ref{prop:restriction}
and exact at $\QCiso{\pi_0(G)}$ by Lemma~\ref{lem:extension}.
Here we show that it is also exact at $\QCiso{G}$.
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Proposition~\ref{prop:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.

>>>>

\begin{abstract}
Motivated by an application to the geometrization and categorification of quasicharacters of $p$-adic tori,
this paper extends Deligne's function--sheaf dictionary for characters of connected commutative algebraic groups over finite fields to characters of smooth commutative group schemes over finite fields.
 \end{abstract}

@@

@@

As indicated above, our interest in the function--sheaf dictionary for smooth commutative group schemes over finite fields comes from an application to $p$-adic representation theory. 
We found, however, that this application applies to all local fields $K$ with finite residue field $\Fq$ and all connected commutative algebraic groups $X_K$ over $K$ that do not contain a subgroup of type $\mathbb{G}_\text{a}$, such as Abelian varieties, unipotent $K$-wound groups, and of course, algebraic tori. 
Such algebraic groups $X_K$ always admit a Neron model $X$ over $R$, the ring of integers of $K$.
In this paper we show that quasicharacters of $X_K$ are geometrized and categorified by quasicharacter sheaves on the
Greenberg transform $\Gr_R(X)$ of the N\'eron model $X$; although not locally of finite type, $\Gr_R(X)$ is a commutative group scheme over $\Fq$ and also a projective limit of smooth commutative group schemes $\Gr^R_n(X)$.
This allows us to adapt our work on quasicharacter sheaves to construct a category $\QC(X_K)$ of quasicharacter sheaves for $X_K$, which are certain sheaves on $\Gr_R(X)\times_{\Spec{\Fq}} \Spec{\bFq}$, with extra structure.
The ability to generalize the function--sheaf dictionary to non-connected group schemes, of which the component group of the special fibre of the Néron model $X$ is an important example, plays a crucial role in this application
since these group schemes are not connected in general.

Specializing to the case when $X_K$ is an algebraic tori $T$ over $K$, and writing $\underline{T}$ for the Neron model $X$ in this case, this paper gives a canonical short exact sequence 
\begin{equation}
0 \to \Hh^2(X_*(T)_{\mathcal{I}_K},\EEx)^{\Weil{}} \to \QCiso{T} \to \Hom_\text{qc}(T(K),\EEx) \to 0,
\end{equation}
where $\Hom_\text{qc}(T(K),\EEx)$ denoted the group of quasicharacters of $T(K)$.
We further show that automorphism groups in the $\QC(\Gr_R(\underline{T}))$ are given, for every quasicharacter sheaf $\qcs{F}$ on $\Gr_R(X)$, by
\begin{equation}
\Aut(\qcs{F}) \iso (\check{T}_\ell)^{\Weil{K}}
\end{equation}
where $\Weil{K}$ is the Weil group for $K$ and $\check{T}_\ell$ is the $\ell$-adic dual torus to $T$.
%We also use quasicharacter sheaves to categorify the transfer of quasicharacters of tori between characteristic $0$ and characteristic $p$ local fields.

By any measure, there are more quasicharacter sheaves for $T$ than quasicharacters of $T(K)$.  
%\todo{Maybe move this from the introduction into the body of the paper somewhere.}
In this regard, we are reminded of the work of David Vogan \cite{vogan:93a}, in which he finds a geometrization of complete Langlands parameters for $p$-adic groups as equivariant perverse sheaves on Vogan varieties and, in the process, is lead to study the representations of all the pure rational forms of the $p$-adic group simultaneously.
A similar phenomenon appears in recent work by Joseph Bernstein in which his geometric Ansatz leads to the study of certain sheaves on the stacky classifying space of the $p$-adic group, resulting in a category which appears to be tied to the representations of all the pure rational forms of the $p$-adic group.
Indeed, he has suggested to us that our category of quasicharacters for $T$ may be tied to characters of all the pure rational forms of $T$.

@@

%\begin{acknowledgements}
\subsection{Acknowledgments}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We thank Takashi Suzuki for some very helpful observations and suggestions, including those related to our use of the Hochschild-Serre spectral sequence.
We thank Joseph Bernstein for suggesting a connection between our category of quasicharacter sheaves on $G$ and his sheaves on the stacky classifying space of $G$.
Finally, we gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our week-long stay in May 2012.

@@

@@

We will describe the group $\QCiso{G}$ in Theorem \ref{thm:snake}
and the sets $\Hom(\qcs{L},\qcs{L}')$ in Theorem \ref{thm:autornaught}; in this way we provide a complete description of the category $\QC(G)$.
%
In the meantime, we close this section with an elementary observation about $\Hom(\qcs{L},\qcs{L}')$.

\begin{lemma}\label{lem:autornaught}
Let $G$ be a smooth commutative group scheme over $\Fq$.
If $\qcs{L}$ and $\qcs{L}'$ are quasicharacter sheaves on $G$ then
every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism. 
\end{lemma}

\begin{proof}
Suppose $\rho \in \Hom(\qcs{L},\qcs{L}')$.
We prove the lemma by considering the linear transformations $\rho_{\bar g} : \gqcs{L}_{\bar g} \to \gqcs{L}_{\bar g}$ at the stalks above geometric points ${\bar g}$ on $G$ and showing that, either each $\rho_{\bar g}$ trivial or each $\rho_{\bar g}$ is an isomorphism.
(This idea is expanded upon in Section~\ref{sec:stalks}.)
Let ${\bar e}$ be the geometric point above the identity $e$ for $G$ determined by our choice of algebraic closure $\bFq$ of $\Fq$.
If $\rho_{\bar e} = 0$ then the second diagram in \ref{QC.4} implies that $\rho_{\bar g} = 0$ for all ${\bar g}$, in which case $\rho$ is trivial.
On the other hand, if $\rho_{\bar e}$ is nontrivial then the second diagram in \ref{QC.4}  implies that $\rho_{\bar g}$ is nontrivial for all ${\bar g}$ and thus an isomorphism, since the stalk of quasicharacter sheaves are one-dimensional; in this case $\rho$ is an isomorphism.
\end{proof}

@@

Let $(\gqcs{L},\phi)$ be a Weil sheaf on $G$. Every $g\in G(\Fq)$
determines a point geometric point $\bg$ fixed by $\Frob{G}$. 
Together with the canonical isomorphism $(\Frob{G}^*\gqcs{L})_{\bg} \iso  \gqcs{L}_{\Frob{G}(\bg)}$, in this way the automorphism $\phi$ determines an automorphism $\phi_{\bg}$ of the $\EE$-vector space $\gqcs{L}_{\bg}$; let $\trace(\phi_{\bg};\gqcs{L}_{\bg})$ be the trace of $\phi_{\bg} \in \Aut_{\EE}(\gqcs{L}_{\bg})$ and let $\trFrob{(\gqcs{L},\phi)} : G(\Fq)\to \EE$ be the function defined by 
\begin{equation}\label{trWeil}
%\begin{equation}\label{trWeil}
\trFrob{(\gqcs{L},\phi)}(g) \ceq \trace(\phi_{\bg};\gqcs{L}_{\bg}),
%\end{equation}
\end{equation}
commonly called the {\em trace of Frobenius of $(\gqcs{L},\phi)$}.
Note that if $(\gqcs{L},\phi) \iso (\gqcs{L'},\phi')$ as Weil sheaves, 
then $\trFrob{(\gqcs{L},\phi)} = \trFrob{(\gqcs{L'},\phi')}$ as functions on $G(\Fq)$.

@@

\begin{lemma}\label{lem:pullback}
  If $f : H\to G$ is a morphism of smooth commutative group schemes over $\Fq$ then
  \begin{align*}
  f^* : \QC(G) &\to \QC(H) \\
  (\gqcs{L},\mu,\phi) &\mapsto (\bf^*\gqcs{L},(\bfxf)^*\mu,\bf^*F)
  \end{align*}
  defines a monoidal functor dual to $f \colon H(\Fq) \to G(\Fq)$:
  \[
  \begin{tikzcd}[row sep=20, column sep=30]
   \QCiso{G} \rar{f^*} \arrow[swap]{d}{\TrFrob{G}} & \QCiso{H} \dar{\TrFrob{H}} \\
   G(\Fq)^* \rar & H(\Fq)^*
  \end{tikzcd}
  \]
  is a commutative diagram of groups.  Moreover, $(f\circ g)^* = g^* \circ f^*$.
\end{lemma}
\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. 
  Pullback by $\bf$ takes rank~one local systems to rank~one local systems.
  To see that $(\bfxf)^* \mu$ satisfies \ref{QC.2},
  apply the functor $(\bfxf)^*$
  to \ref{QC.2} for $\qcs{L}$ and use the canonical isomorphism
  $(\bfxf)^*(\LxL) \iso \bf^*\gqcs{L} \tight{-3}{\boxtimes}{-3} \bf^*\gqcs{L}$.
  To show that $f^*\qcs{L}$ satisfies
  \ref{QC.3}, apply the same functor to \ref{QC.3} for $\qcs{L}$.
  Since $f$ is a morphism of group schemes defined over $\Fq$
  it provides isomorphisms $(\bfxf)^*\Frob{\GxxG}^* \iso \Frob{\GxxG}^* (\bfxf)^*$
  and $(\bfxf)^* \bm^*\iso \bm^* \bf^*$ between functors of constructible sheaves.

@@

\begin{lemma}\label{lem:product}
If $G_1$ and $G_2$ are smooth commutative group schemes over $\Fq$ then the rule
$\boxtimes : (\qcs{L}_1,\qcs{L}_2)\to \qcs{L}_1\boxtimes\qcs{L}_2$ defines an equivalence of categories
\[
\QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)
\]
such that
\[
\begin{tikzcd}[column sep=60]
\arrow{d}{\TrFrob{G_1} \times \TrFrob{G_2}} \QCiso{G_1}\times \QCiso{G_2} \arrow{r}{\boxtimes}
& \arrow{d}{\TrFrob{G_1\times G_2}} \QCiso{G_1\times G_2}\\
(G_1)(\Fq)^*\times (G_2)(\Fq)^* \arrow{r}{(\chi_1,\chi_2)\mapsto \chi_1\otimes \chi_2}  & (G_1\times G_2)(\Fq)^*
\end{tikzcd}
\]
commutes.
\end{lemma}
\begin{proof}
The only non-trivial part of the proof is to show that $\boxtimes : \QC(G_1)\times \QC(G_2) \to \QC(G_1\times G_2)$
is essentially surjective, so we will only address that point here.

@@

Now, let $\qcs{L} \ceq (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
Set $\qcs{L}_1 \ceq q_1^* \qcs{L}$ and $\qcs{L}_2 \ceq q_2^* \qcs{L}$;
by Lemma~\ref{lem:pullback}, $\qcs{L}_1$ a quasicharacter sheaf on $G_1$
and $\qcs{L}_2$ is a quasicharacter sheaf on $G_2$.
We will obtain an isomorphism $\qcs{L} \iso  \qcs{L}_1\boxtimes \qcs{L}_2$.
Applying the functor $f^*$ to the isomorphism $\mu$ yields
\begin{equation}\label{eq:fm}
f^*\mu : f^* m^* \gqcs{L} \to f^*(\gqcs{L}\boxtimes \gqcs{L}) .
\end{equation}
We have already observed that $m\circ f = \id_G$, so $f^* m^* \gqcs{L} = \gqcs{L}$.
On the other hand, $f^*(\gqcs{L}\boxtimes \gqcs{L}) = f^*p_1^*\gqcs{L}\otimes f^* p_2^*\gqcs{L}$
by definition.  Since $f^*p_1^*\gqcs{L} = r_1^* q_1^* \gqcs{L}$ and $f^*p_2^*\gqcs{L} = r_2^* q_2^* \gqcs{L}$
we get that $f^*(\gqcs{L}\boxtimes \gqcs{L}) = \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It follows that \eqref{eq:fm} gives an isomorphism $\gqcs{L} \to  \gqcs{L}_1\boxtimes \gqcs{L}_2$.
It is a matter of routine to show that this morphism satisfies
\ref{QC.3} as it applies here,
from which it follows that we have exhibited an isomorphism
$\qcs{L} \to \qcs{L}_1\boxtimes \qcs{L}_2$ of quasicharacters sheaves on $G\times G$.
This completes the proof that $\boxtimes$ is essentially surjective.
\end{proof}

@@

\begin{proof}
The first part of Lemma~\ref{lem:pullback} shows that $F_1$ is a functor,
while the second part shows that Trace of Frobenius is a natural transformation
$T: F_1 \to F_2$. When further combined with Lemma~\ref{lem:product},
we see that $F_1$ is an additive functor and $T: F_1 \to F_2$ is a natural
transformation between additive functors,
concluding the proof of Proposition~\ref{prop:functorialG}.
\end{proof}

@@

\subsection{Stalks of quasicharacter sheaves}\label{sec:stalks}

@@

\begin{proposition}
\begin{enumerate}
\item The second page is given by $E_2^{i,j} = \Hh^i(\Weil{}, \Hh^j(G(\bFq), \EEx))$,
\item there is an isomorphism $\Hh^n(\Weil{} \ltimes G(\bFq), \EEx) \cong \Hh^n(E_G^\bullet)$,
\item there is a filtration $\Hh^n(\Weil{} \ltimes G(\bFq), \EEx) = F_n \supset \cdots \supset F_0 \supset F_{-1} = 0$ where $F_i / F_{i-1} \cong E_{\infty}^{i, n-i}$.
\end{enumerate}
\end{proposition}
%\todo{Is this a lemma? Is the corollary a proposition?}

@@

\begin{proof}
We have already seen, in Lemma~\ref{lem:autornaught}, that every $\rho\in \Hom(\qcs{L},\qcs{L}')$ is either trivial or an isomorphism.
Now suppose $\rho \in \Aut(\qcs{L})$.
The second diagram in \ref{qc.4} shows that the association $x \mapsto \rho_x$ is a homomorphism from $G(\bFq)$ to $\EEx$ and the first diagram in \ref{qc.4} shows that it factors through $G(\bFq)_{\Weil{}}$.  Conversely, if $\rho : G(\bFq)_{\Weil{}} \to \EEx$ is any homomorphism then defining $\rho_x$ as multiplication by $\rho(x)$ will satisfy the two diagrams.  Composition of morphisms corresponds to pointwise multiplication in this correspondence, showing that the resulting bijection is actually a group isomorphism.
\end{proof}

@@

\begin{proposition}\label{prop:connected}
 If $G$ is a connected, commutative algebraic group over $\Fq$ then:
 \begin{enumerate}
 \item $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is an isomorphism of groups;
 \item every quasicharacter sheaf on $G$ is isomorphic to one defined by a discrete isogeny;%as in Proposition~\ref{prop:finite};
 \item $\Aut(\qcs{L}) = 1$, for all quasicharacter sheaves $\qcs{L}$ on $G$;
%  \item for all quasicharacter sheaves $\qcs{L}$, $\qcs{L}'$ on $G$ and for every $\rho \in \Hom(\qcs{L},\qcs{L}')$, either $\rho$ is trivial or an isomorphism.
 \end{enumerate}
 \end{proposition}

\begin{proof}
In Lemma~\ref{lem:bounded_connected} we saw that every quasicharacter sheaf $\qcs{L}$ on $G$ descends to $G$; let $\qcs{E}$ be an object in $\bQC(G)$ for which $B_G(\qcs{E}) \iso \qcs{L}$.
 Since the functor $B_G : \bQC(G) \to \QC(G)$ is full and faithful, $\Aut(\qcs{L}) = \Aut(\qcs{E})$.
From here, Deligne's function--sheaf dictionary for connected commutative algebraic groups over finite fields \citelist{\cite{deligne:SGA4.5}*{Sommes trig.}, \cite{laumon:87a}*{1.1.3}}  gives us all we need, as we briefly recall.
The key observation is that, by étale descent, pullback by the Lang isogeny $\Lang : G\to G$ defines an equivalence of categories from $\bQC(G)$ to the category of $1$-dimensional $\ell$-adic representations $V$ of $G(\Fq)$ equipped with an isomorphism $V \to V\otimes V$ such that the trace of $\Lang^*\qcs{E}$ is $\trFrob{\qcs{E}}^{-1}$ for every $\qcs{E}$ in $\bQC(G)$. 
Using the equivalence given by $\Lang^*$, if $\Lang^*\qcs{E} = V$ with character $\psi$ and isomorphism $V\to V\otimes V$ then 
$\qcs{E} \iso (\Lang_! V)_\psi$, the $\psi$-isotypic component of the push-forward $\Lang_* V$ ($\Lang$ is proper, so $\Lang_* = \Lang_!$) of the constant sheaf $V$ on $G$ equipped with an isomorphism $\mu_\qcs{E} : m^* \qcs{E} \to \qcs{E} \boxtimes\qcs{E}$ determined by $V\to V\otimes V$. 
Moreover, $\Aut(\qcs{E})$ is the group of automorphisms of $\rho : V\to V$ for which 
\[
\begin{tikzcd}
\arrow{d}{} V \arrow{r}{\rho} & V\arrow{d}{}\\
V\otimes V \arrow{r}{\rho\otimes \rho} & V\otimes V
\end{tikzcd}
\]
commutes. 
Since the only such isomorphism $\rho$ is $\id_V$, it follows that $\Aut(\qcs{E}) = 1$, completing the proof.
\end{proof}

@@

\section{The function--sheaf dictionary for smooth commutative group schemes}\label{sec:main}

@@

\begin{proof}
The sequence is exact at $\QCiso{G^0}$ by Proposition~\ref{prop:restriction}
and exact at $\QCiso{\pi_0(G)}$ by Lemma~\ref{lem:extension}.
Here we show that it is also exact at $\QCiso{G}$.
First note that $\iota_0^* \circ \pi_0^*$ is trivial by Lemma~\ref{lem:pullback}.
So it suffices to show that if $\qcs{L} = (\gqcs{L},\mu,\phi)$ is a quasicharacter sheaf on $G$
with $\qcs{L}\vert_{G^0} = (\EE)_{G^0}$ then $\qcs{L}$ is in the essential image of $\pi_0^*$.
